---
title: "Differential geometry"
output: html_notebook
---

```{r}
library(plot3D)
library(magrittr) # pipe: %>%
library(zeallot) # unpacking: %<-% 

## very interesting!!!:
# relist(flesh, skeleton)
# unlist


# pointwiseProduct(a=1:3,b=1:3)
# [1] 1 4 9
pointwiseProduct <- function(...) Reduce(`*`, list(...))


# dotProduct(1:3,1:3)
# [1] 14
dotProduct <- function(x, y) sum(x*y)
# pointwiseDotProduct <- function(l1, l2) 0


# listToMatrix(list(x=1:3,y=1:3,z=1:3)) 
#   [,1] [,2] [,3]
# x    1    2    3
# y    1    2    3
# z    1    2    3
listToMatrix <- function(list) do.call(rbind, list)
  

# pointwiseNorm(x=1:3,y=1:3,z=1:3) 
# [1] 1.732051 3.464102 5.196152
pointwiseNorm <- function(...) {
  sapply(list(...), function(v) v^2) %>%  rowSums %>% sqrt
}


# normOfVectors(x = 1:3, y = 1:3, z = 1:3)
# [1] 6.480741
# ((1:3)^2+(1:3)^2+(1:3)^2) %>% sum %>% sqrt
# [1] 6.480741
normOfVectors <- function(...) cbind(...) %>% norm(., "F")


# indexOf.FromToLength(from = -10, to = 20, length.out = 10, e = 5)
# [1] 6
indexOf.FromToLength <- function(from, to, length.out, e, lower = TRUE) {
  ds <- (to - from)/(length.out-1)
  i <- (e - from)/ds + 1
  round(i)
}


# indexOf.FromToBy(from = -10, to = 20, by = 31/10, e = 5)
# [1] 6
indexOf.FromToBy <- function(from, to, by, e, lower = TRUE) {
  i <- (e - from)/by + 1
  round(i)
}


# seq.by(seq(-10, 20, length.out = 10))
# seq.by(seq(-10, 20, length.out = 10))
# [1] 3.333333
# -10 + 9 * 3.333333
# [1] 20
seq.by <- function(seq) (seq[length(seq)] - seq[1]) / (length(seq) - 1)


# indexOf.Sequence(seq = seq(-10, 20, length.out = 10), e = 5)
# [1] 6
indexOf.Sequence <- function(seq, e, lower = TRUE) {
  len <- length(seq)
  ds <- (seq[len] - seq[1])/(len-1)
  i <- (e - seq[1])/ds + 1
  round(i)
}


```

$$
\newcommand{\bra}[1]{\left\langle#1\right|}
\newcommand{\ket}[1]{\left|#1\right\rangle}
\newcommand{\bracket}[2]{\left\langle#1 \vphantom{#2}\middle|  #2 \vphantom{#1}\right\rangle}
\newcommand{\matrixel}[3]{\left\langle #1 \vphantom{#2#3} \right| #2 \left| #3 \vphantom{#1#2} \right\rangle}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
$$
  

```{r}
(function(){
  # approxfun(1,1)
  # splinefun(1,1)
  
})()
```



## Parametric curves


Parametric curve is somthing like this:


$$
\alpha : I \rightarrow R^3\\
\alpha(t) = x(t)\ket{1} + y(t)\ket{2} + z(t)\ket{3}
$$
It is function mapping interval I into $R^3$ vector.

Norm of curve value, is length from origin to curve point at t:

$$
\begin{align}
\norm{\alpha(t)} &= \sqrt{x^2(t) + y^2(t) + z^2(t)}\\
&= \sqrt{\alpha(t) \cdot \alpha(t)}
\end{align}
$$



```{r}

(function(){

  op <- par(
    mfrow = c(1,2)
    # mai = c(0.4,0.4,0.1,0.1),
    # tck = 0.05 # not working for 3D, should specify somewhere else...
  )
  
  koniczyna <- function(t, s = 1) list(x = (t^3 - 4*t)/s, y= (t^2 - 4)/s, z = 0*t+1)
  c(x,y,z) %<-% koniczyna(seq(-2.5,2.5, length.out = 100))
  
  
  lines3D (
    x, y, z, 
    col = "red", cex.axis = 0.8, cex.lab = 1.2,
    # add = TRUE,
    # xlim = c(-2,2),ylim=c(-2,2),cex.lab=1, 
    r = 5, # eyepoint dist from center
    d = 3, # perspective greaten < 1 < prespective lessen
    main = "Koniczyna",
    ticktype = "detailed", colkey = FALSE, bty = "b2")
  
  
  cylinder <- function(t, a = 1, b = 1) list(x = a*cos(t), y = a*sin(t), z = b*t)
  c(x,y,z) %<-% cylinder(seq(0, 20, length.out = 100))
  
  lines3D (
    x, y, z, 
    col = "green", cex.axis = 0.8, cex.lab = 1.2,
    # add = TRUE,
    xlim = c(-2,2),ylim=c(-2,2),
    r = 5, # eyepoint dist from center
    d = 3, # perspective greaten < 1 < prespective lessen
    main = "Cylinder",
    ticktype = "detailed", colkey = FALSE, bty = "b2")
  
  par(op)

})()

```




### Example: Attractor??


Let α(t) be a parametrized curve which does not pass through the origin.
If α(t0) is a point of the trace of α closest to the origin and α′(t0) = 0, show
that the position vector α(t0) is orthogonal to α′(t0).


```{r}

(function(){
curv <- function(t) list(x=t^2+1, y=t-1)
dCurv <- function(t) list(x=2*t, y=0*t+1)
t <- seq(-2,2, length.out = 50)
c(x,y) %<-% curv(t)
c(dx,dy) %<-% dCurv(t)


curv2 <- function(t) sqrt((t^2+1)^2 + (t-1)^2)
t0 <- optimize(curv2, c(-2,2))$min

## xyPointToStr(list(x=1.1,y=2.22)) == (1.1,2.2)
xyPointToStr <- function(point, digits = 2) {
  paste0(
    "(",
    signif(point$x, digits = digits), 
    ",",
    signif(point$y, digits = digits),
    ")")
}

## tToXyToStr(0.55, list(x=1.1,y=2.22)) == "0.5 -> (1.1,2.2)"
tToXyToStr <- function(t, xy, digits = 2) paste0(signif(t, digits = 2), " -> ", xyPointToStr(xy))

plot(range(x,dx),range(y,dy), 
     type = "n", # dont draw data
     xlab = "x",
     ylab = "y"
     # ann = FALSE,
     # xaxt = "n", # no X axis ticks and ticks labels
     # yaxt = "n"  # no Y axis ticks and ticks labels
     )
# par(new =  TRUE) # drows on existing plot, it is reset after plot call


lines(x,y, col = "dark blue", lwd = 2, lty = 1)
points(x,y, col = "dark blue")
lines(dx,dy, type = "l", col = "dark red", lwd = 2, lty = 2)
points(dx,dy, col = "dark red")
arrows(0,0, curv(t0)$x, curv(t0)$y, col = "dark blue")
text((curv(t0)$x - 0)/2, (curv(t0)$y - 0)/2, 
     paste0(signif(t0, digits = 2), " -> ", xyPointToStr(curv(t0))),
     col = "dark blue")
arrows(0,0, dCurv(t0)$x, dCurv(t0)$y, col = "dark red")
text((dCurv(t0)$x - 0)/2, (dCurv(t0)$y - 0)/2, 
     paste0(signif(t0, digits = 2), " -> ", xyPointToStr(dCurv(t0))),
     col = "dark red")
abline(h=0)
abline(v=0)
grid()

})()
```



## Curves parametrized by arc length

Differential of curve α(t) is α'(t) given by (for 3D):

$$
\alpha'(t) = x'(t)\ket{1} + y'(t)\ket{2} + z'(t)\ket{3}
$$

Arc length s of regular (points t with α'(t) = 0 are singular) parametrized curve α is:

$$
\begin{cases}
s(t) &= \int_{t_0}^{t} \norm{\alpha'(t)} \, dt\\
\norm{\alpha'(t)} &= \sqrt{(x'(t))^2 + (y'(t))^2 + (z'(t))^2}
\end{cases}
$$

||α'(t)|| is tangent curve point distance from origin.


Given arc length of curve we can deduce distance from origin of curve tangent:

$$
\norm{\alpha'(t)} = \frac{ds(t)}{dt}
$$

Co jeśli ||α'(t)|| = const ? Nie oznacza to, że α(t) jest prostą (ponieważ norma stycznych jest stała).
Ale wtedy długość łuku s będzie w skali const odpowiadała parametrowi t:

$$
\begin{cases}
\norm{\alpha'(t)} &= a\\
s(t) &= \int_{t_0}^{t} a \, dt = a(t - t_0)
\end{cases}
$$

Zakładając teraz że $t_0 = 0$ oraz $a = 1$ mamy:

$$s = t$$

Oznacza to, że krzywa α dla parametru t zachowuje się tak samo
jak krzywa α dla parametru s (długości krzywej).
Inaczej mówiąc, punkt krzywej α po przebyciu drogi s będzie
taki sam jak punkt krzywej α po upływie czasu t, jeżeli
wartości s i t będą sobie równe.


Ciekawym spostrzerzeniem jest to, że mając dowolnie sparametryzowaną krzywą α
możemy ją zawsze przedstawić w postaci sparametryzowanej przez jej własną długość s.
Dzieje się tak ponieważ istnieje oczywisty związek pomiędzy 
długością krzywej a jej zapisem parametrycznym. Robimy to następująco:


[Integral of square root of sin](https://math.stackexchange.com/questions/1469846/integration-of-sqrt-sin-x-dx)

$$
\begin{cases}
u=\sqrt{\sin x}\\
x=\sin^{-1}u^2\\
dx=\dfrac{2u}{\sqrt{1-u^4}}du\\
\end{cases}
$$

Therefore:


$$
\therefore\int\sqrt{\sin x}~dx =\int\dfrac{2u^2}{\sqrt{1-u^4}}du
$$


$$
\int\dfrac{2u^2}{\sqrt{1-u^4}}du
=\int2u^2\sum\limits_{n=0}^\infty\dfrac{(2n)!u^{4n}}{4^n(n!)^2}du\\
=\int\sum\limits_{n=0}^\infty\dfrac{(2n)!u^{4n+2}}{2^{2n-1}(n!)^2}du
=\sum\limits_{n=0}^\infty\dfrac{(2n)!u^{4n+3}}{2^{2n-1}(n!)^2(4n+3)}+C
$$

Reasuming:

$$
\int\sqrt{\sin x}~dx
=\sum\limits_{n=0}^\infty\dfrac{(2n)!\sin^{2n+\frac{3}{2}}x}{2^{2n-1}(n!)^2(4n+3)}+C
$$



$$
\begin{cases}
u=\sqrt{\cos x}\\
x=\cos^{-1}u^2\\
\dfrac{d}{dx}\cos^{-1}(x) = - \dfrac{1}{\sqrt{1-x^2}}\\
dx=-\dfrac{2u}{\sqrt{1-u^4}}du\\
\end{cases}
$$

$$
\int\sqrt{2 - 2\cos x}~dx\\
u=\sqrt{2 - 2\cos x}\\
x=\cos^{-1}u^2\\
$$





## Local theory of curves




### Example: tangent, curvature and torsion


Helix have parametric equation:

$$
\begin{cases}
\alpha(s) = \left( a \cos(\frac{s}{c}), a \sin(\frac{s}{c}), b\frac{s}{c}  \right)\\
c^2 = a^2 + b^2
\end{cases}
$$

Parameter s is arc length iff norm of α' equals 1. Helix tangent is:

$$
\alpha'(s) = \left( -\frac{a}{c} \sin(\frac{s}{c}), \frac{a}{c} \cos(\frac{s}{c}), \frac{b}{c}  \right)
$$

Norm of helix tangent is:

$$
\begin{align}
\alpha'(s)\alpha'(s) &= \frac{a^2}{c^2}\sin^2(\frac{s}{c}) + \frac{a^2}{c^2}\cos^2(\frac{s}{c}) + \frac{b^2}{c^2}\\
&= \frac{a^2}{c^2} + \frac{b^2}{c^2} = \frac{a^2 + b^2}{c^2} = 1
\end{align}
$$

It can be seen from figure below.


```{r}

(function(){
  
  helixGen <- function(a, b) {
    force(a)
    force(b)
    c <- sqrt(a^2+b^2)
    function(arcLen) list(
      x = a*cos(arcLen/c),
      y = a*sin(arcLen/c),
      z = b*arcLen/c
    )
  }
  
  helixTangentGen <- function(a,b) {
    c <- sqrt(a^2+b^2)
    function(t) list(
      dx = -a/c*sin(t/c),
      dy = a/c*cos(t/c),
      dz = b/c + 0*t
    )
  }
  
  
  a <- 0.5
  b <- 1
  helix <- helixGen(a=a,b=b)
  helixTangent <- helixTangentGen(a=a,b=b)
  arcLen = seq(from=0, to=20, length.out = 50)
  # arcLen = seq(from=0, to=20, by = 1)
  arcLenDiff = seq.by(arcLen)
  c(x,y,z) %<-% helix(arcLen)
  c(dx,dy,dz) %<-% helixTangent(arcLen)
  
  
  
  lines3D (
    x, y, z, 
    col = "red", cex.axis = 0.8, cex.lab = 1.2,
    # xlim = c(-2,2),ylim=c(-2,2),cex.lab=1, 
    r = 5, # eyepoint dist from center
    d = 3, # perspective greaten < 1 < prespective lessen
    main = paste0("Helix with parameters ","a = ", a, ", b = ", b),
    ticktype = "detailed", colkey = FALSE, bty = "b2")
 
  
  lines3D (
    dx, dy, dz, 
    col = "blue", cex.axis = 0.8, cex.lab = 1.2,
    # xlim = c(-2,2),ylim=c(-2,2),cex.lab=1, 
    r = 5, # eyepoint dist from center
    d = 3, # perspective greaten < 1 < prespective lessen
    main = paste0("Tangent to helix with parameters ","a = ", a, ", b = ", b),
    ticktype = "detailed", colkey = FALSE, bty = "b2")
  
  
  plot(arcLen, pointwiseNorm(dx,dy,dz), 
       main = paste0("Norm of tangent to helix with parameters ","a = ", a, ", b = ", b), 
       type = "l")
  
  
  pointwiseNorm(dx,dy,dz)[3] %>% print
  
  # diff expect dt = 1
  # if it is some thing other, result should be rescaled
  # diff with differences = 2 it is probably newton-cotes formula of order 2, need check
  pointwiseNorm(diff(x)/arcLenDiff,diff(y)/arcLenDiff,diff(z)/arcLenDiff)[2] %>% print
})()

```




```{r}

(function(){
  
  
})()



```



































